---
title: "Coursera Project 1"
output: html_document
---

First we read our data from the working directory into R.
```{r}
dta<-read.csv("~/repdata-data-activity/activity.csv")
```
###Calculate mean number of steps taken per day
First we can calculate the total number of steps taken per day and plot this as a histogram
```{r}
totsteps<-tapply(dta$steps, dta$date, sum)
hist(totsteps)
```

Now we take the **mean** and **median**of the steps taken per day, removing any NA values
```{r}
meansteps<-tapply(dta$steps, dta$date, mean, na.rm=TRUE)
mediansteps<-tapply(dta$steps, dta$date, mean, na.rm=TRUE)
meansteps
mediansteps
```

###Average daily activity pattern
To make a plot of the average number of steps for each 5-minute interval across all days we first take the average for each interval and then plot this. this shows the interval with the highest average steps
```{r}
avsteps<-aggregate(dta$steps, list(dta$interval), mean, na.rm=TRUE)
plot(avsteps[,1], avsteps[,2], type='l', xlab="5-minute intervals", ylab="average steps")
```

This shows us that the maximum average steps takes place at ***interval 835***

###Inputting missing values
First we count the number of missing values in the whole dataset.
After this we replace all NA values with the mean value of their respective 5-minute interval.
```{r}
sum(is.na(dta))
dta2<-dta
for (i in 1:nrow(dta2)){
if (is.na(dta2$steps[i])){
    dta2$steps[i] <- avsteps[which(dta2$interval[i]==avsteps$Group.1), 2]
      }
}
head(dta2)
sum(is.na(dta2))
```
Next we make a histogram of the total steps taken per day.
```{r}
totsteps2<-tapply(dta2$steps, dta2$date, sum)
hist(totsteps2)
meansteps2<-tapply(dta2$steps, dta2$date, mean, na.rm=TRUE)
mediansteps2<-tapply(dta2$steps, dta2$date, mean, na.rm=TRUE)
meansteps2
mediansteps2
```
Imputing these NA values slightly alters the distribution, making it a little flatter. The peak frequency remains in the same place, however. The mean and median values for steps taken per day only change on those days that had NA values.

###Finding differences between weekends and weekdays
First convert the date variable to date type, and create a new variable that tells us whether each date is a weekend or weekday
```{r}
dta2$date<-as.Date(dta2$date, format="%Y-%m-%d")
dta2$day<-weekdays(dta2$date)
dta2$week<-c()
for (i in 1:nrow(dta2))
  if(dta2$day[i]=="Saturday"||dta2$day[i]=="Sunday"){
  dta2$week[i]<- "Weekend"
  } else{
  dta2$week[i]<- "Weekday"
  }
dta2$week<-factor(dta2$week)
```

Now we can plot the average number of steps taken for each 5-minute interval across weekend and weekdays.
```{r}
avsteps2<-aggregate(dta$steps, list(dta2$interval, dta2$week), mean, na.rm=TRUE)
avsteps2$Group.2<-factor(avsteps2$Group.2)
par(mfrow=c(2,1))
plot(avsteps2$Group.1[avsteps2$Group.2=="Weekday"], avsteps2$x[avsteps2$Group.2=="Weekday"], type='l', ylab="average steps", main="Weekday Activity", xlab="")
plot(avsteps2$Group.1[avsteps2$Group.2=="Weekend"], avsteps2$x[avsteps2$Group.2=="Weekend"], type='l', xlab="5-minute intervals", ylab="average steps", main="Weekend Activity")
```

So we can see that during weekends activity is on average higher later in the day.
